---
title: "MoodMapper"
author: "Richa Jain"
format: html
---
```{r}
library(dplyr)
library(tidyverse)
library(readr)
library(tidytext)
```

Resource used: https://bookdown.org/valerie_hase/TextasData_HS2021/tutorial-11-preprocessing.html

## Read in dataset

```{r}
mood_data <- read.table("data/val.txt", sep = ";", header = FALSE)
colnames(mood_data) <- c("Description", "Mood")
mood_data
```

## Text Preprocessing

1. Check for encoding issues: 
```{r}
mood_data$Description[1]
```
```{r}
writeLines(mood_data$Description[1])
```

- reading the text seems to have worked without any encoding issues. 

2. Tokenization:
- remove numbers, punctuation marks, urls
- remove features that provide little informational value
- convert all text to lowercase in case not already lowercase
```{r}
library(quanteda)
tokens <- tokens(mood_data$Description, what = "word",
                 remove_punct = TRUE,
                 remove_numbers = TRUE,
                 remove_url = TRUE)
tokens <- tokens_tolower(tokens)
```

3. Remove stopwords
- tidytext has a database of commonly used stop words
```{r}
stop_words
```

```{r}
tokens <- tokens_remove(tokens, stopwords("english"))
tokens <- tokens_wordstem(tokens)
```
























